Writing Trash Bin

### 2024-03-19 paper.qmd
finished perceptual task section

It is difficult to find examples of uncertainty visualisation experiments where the plots do contain the same information, however when they do, the results align with existing information visualisation research.

*Other examples* (*Cite Gap 15*).

If plots that have different information are compared, whichever directly depicts the ground truth statistic (or has information that can easily be convered to that ground truth statistic) will perform the best. If plots that have the same information are compared, the plot that requires a perceptual task that is higher on the hierarchy will perform the best. Therefore, when it comes to any task that involves value abstraction, many uncertainty visualisation evaluation studies are simple repeating already established results in statistics or visualisation.

Therefore, these explanations for the difference fall short of explaining *why* uncertainty should be treated any differently than any other variable.  The psychological effects of uncertainty are only relevant for decision making and have nothing to do with value abstraction, and the information overload issue is not specific to uncertainty and is simply an example of a larger dimensionality issue.

Despite these results, it is clear that authors intuitively *feel* that uncertainty is different even though this difference has not been apparent in existing uncertainty visualisation experiments. @Kinkeldey2014 also noticed that it is not clear whether or not uncertainty is just another variable, as many value extraction papers treats it as such in their experimental design and uncertainty may need to be in a variable class of its own. @Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty, but did not explain *why* uncertainty information faces more difficulty. After all, confidence intervals, minimums, maxmimums, variance, and any other statistic that can be considered related to "uncertainty" is just that, a statisitc. There is nothing special about these statistics that should differentiate them from a mean or median and imply their visualisation should be "special" in a way that would warrant its own field. Especially because it seems as though (at least in the case of visual inference) they *are* no different from any other value. So, if we cannot see any difference to uncertainty information when it comes to value abstraction, *when* is uncertainty information different?

# Uncertainty as noise and Uncertainty as signal
In this section we will discuss how *most* attempts to measure the effects of uncertainty are either value extraction experiments with an extra step, or are measuring uncertainty using a proxy that does not actually represent uncertainty. In this section we will talk about the issues of trying to measure uncertainty, the real purpose of uncertainty visualisation, and suggestions in how we can better measure uncertainty in our visualisation experiments.

@@

## The current summaries of findings
## Uncertainty versus Risk
Uncertainty is noise. This may be an obvious statement, but what is not obvious is that the existing literature sees noise as a signal. When we directly ask "What is the varinance of this variable" the variance *becomes* the signal. Noise and signal are not an inherrent property of variance, or meta-data or the aspects of an analysis that are usually secondary considerations, it is a property of the data that *promotes* a message we are trying to infer and the data that *supresses* it. The replication problem in uncertainty visualisation, where experiments are simply repeating established visualisation work, comes from the misunderstandings established in previous sections. Uncertainty is poorly defined, therefore authors do not understand it is not an inherrent property of the data and is actually related to a particular inference question, this incorrect view that uncertainty is always variance or probabilities leads to the class of "uncertainty visualisations" that depict different mathematical objects or visual features with little in the sense of consistent rules, this complicated class has caused visualisation authors to miss that they are simply repeating the established evaluation results from normal information visualisations. This situation is a bit of a mess. It also highlights a unique and facinating problem faced by uncertainty visualisation. If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?

@@

Risk and uncertainty are not only different in how they are communicated, but also why they are communicated. Risk communication focuses on communicating probability around events. These events are typically a binary event that is known to be random, so a version of communication that does not mention risks (e.g. telling someone they will or will not get cancer instead of communicating their risks about it) inherently carries with it a lack of transparency. Additionally risk communication is often communicated to avoid or information about an unfavorable outcome, so framing can be important if we want to influence decisions in a certainty way. These motivations do not extend to the communication of uncertainty information more broadly. While uncertainty information *is* often used to improve decisions, it seems the main motivation in its communication is to offer transparency to the viewer about a statistical analysis. That is, it acts as a form of statistical "hedging" for signals found in an analysis. Interviews with experts in statistic back up this primary motivation, as ignoring uncertainty information of often expressed as being similar to fraud or lying and the goal of "improving decisions" is only seen as a secondary outcome of an appropriately hedged signal [@Hullman2020a; @Manski2020]. This concept of uncertainty means that the work on communicating risk does not cover the field of uncertainty communication as a whole. It covers the specific case where the signal itself *is* uncertain, but this work does not naturally extend to the goal of hedging a signal (i.e. when we want to communicate uncertainty as noise).

>>>>

It is difficult to find examples of uncertainty visualisation experiments where the plots do contain the same information, however when they do, the results align with existing information visualisation research. Technically, a PDF and a mean with confidence intervals both have enough information to extract the mean of the distribution, however they both have a very different computational cost. To extract the mean using a PDF, a participant would need to identify the point along the x axis that splits the area under the curve in half. If a participant is provided with a mean with a confidence interval, extracting the average is a simple task of reading the position on an aligned scale. @Ibrekk1987 found that when asking participants for the "best estimate" (which they thought should be interpreted as the mean of the distribution) of a skewed distribution, participants provided the mean when given a mean with confidence intervals and the mode when given a PDF [@Ibrekk1987]. Similar results to this occur over and over again in the uncertainty visualisation literature. @Gschwandtnei2016 found that visualisations where the start time of an interval could literally be read off the plot (error bars, centered error bars, and ambiguation) performed better than the plots (accumulated probability, gradient, and violin) where the start time involved some guesswork because the drop off was gradual. @Cheong2016 found that participants were better at answering questions when they were explicitly given the relevant probability in text rather than when they needed to read it off a map. *Other examples* (*Cite Gap 15*).

These results show that uncertainty is not technically different to any other variable. When trying to anticipate the results of these studies, we can use the same principles of information equivalence and difficulty of relevant mental tasks to understand which plots will outperform others. This leads us to wonder... why is uncertainty visualisation a field at all?

It is clear that authors intuitively *feel* that uncertainty is different even though this difference has not been captured in existing uncertainty visualisation experiments. Some authors cited uncetainty was special because of its psychological qualities, but we have yet to see any evidence of that here. @Kinkeldey2014 also noticed that it is not clear whether or not uncertainty is just another variable, as many value extraction papers treats it as such in their experimental design and uncertainty may need to be in a variable class of its own. @Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty, but did not explain *why* uncertainty information faces more difficulty. After all, confidence intervals, minimums, maxmimums, variance, and any other statistic that can be considered related to "uncertainty" is just that, a statisitc. There is nothing special about these statistics that should differentiate them from a mean or median and imply their visualisation should be "special" in a way that would warrant its own field. Especially when the core take aways from uncertainty evaluation experiments do not seem to differ at all from the established results in information visualisation. So, what is it that makes authors believe uncertainty is different?

# Are you measuing uncertainty as noise or signal?
Evaluation experiments While uncertainty information *is* often used to improve decisions, it seems the main motivation in its communication is to offer transparency to the viewer about a statistical analysis. That is, it acts as a form of statistical "hedging" for signals found in an analysis. Interviews with experts in statistic back up this primary motivation, as ignoring uncertainty information of often expressed as being similar to fraud or lying and the goal of "improving decisions" is only seen as a secondary outcome of an appropriately hedged signal [@Hullman2020a; @Manski2020]. This concept of uncertainty means that the work on communicating risk does not cover the field of uncertainty communication as a whole. It covers the specific case where the signal itself *is* uncertain, but this work does not naturally extend to the goal of hedging a signal (i.e. when we want to communicate uncertainty as noise).

In this section we will discuss how *most* attempts to measure the effects of uncertainty are either value extraction experiments with an extra step, or are measuring uncertainty using a proxy that does not actually represent uncertainty. In this section we will talk about the issues of trying to measure uncertainty, the real purpose of uncertainty visualisation, and suggestions in how we can better measure uncertainty in our visualisation experiments.

@@

## Uncertainty versus Risk
Uncertainty is noise. This may be an obvious statement, but what is not obvious is that the existing literature sees noise as a signal. When we directly ask "What is the varinance of this variable" the variance *becomes* the signal. Noise and signal are not an inherrent property of variance, or meta-data or the aspects of an analysis that are usually secondary considerations, it is a property of the data that *promotes* a message we are trying to infer and the data that *supresses* it. The replication problem in uncertainty visualisation, where experiments are simply repeating established visualisation work, comes from the misunderstandings established in previous sections. Uncertainty is poorly defined, therefore authors do not understand it is not an inherrent property of the data and is actually related to a particular inference question, this incorrect view that uncertainty is always variance or probabilities leads to the class of "uncertainty visualisations" that depict different mathematical objects or visual features with little in the sense of consistent rules, this complicated class has caused visualisation authors to miss that they are simply repeating the established evaluation results from normal information visualisations. This situation is a bit of a mess. It also highlights a unique and facinating problem faced by uncertainty visualisation. If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?

@@

Risk and uncertainty are not only different in how they are communicated, but also why they are communicated. Risk communication focuses on communicating probability around events. These events are typically a binary event that is known to be random, so a version of communication that does not mention risks (e.g. telling someone they will or will not get cancer instead of communicating their risks about it) inherently carries with it a lack of transparency. Additionally risk communication is often communicated to avoid or information about an unfavorable outcome, so framing can be important if we want to influence decisions in a certainty way. These motivations do not extend to the communication of uncertainty information more broadly.

### 2024-03-19 writingdumpster.qmd
finished perceptual task section

# Moved over notes

@@

# Main section outtakes
There is also the common view that uncertainty as a dimensionality problem. This perspective is not quite right and the literature seems to be about 50/50 on whether uncertainty is a psychological or dimensionality problem.  It is important to keep in mind that just because information is in a graphic, that does not mean it will be "seen". The phenomena of inattentional blindness shows that there is no perception without attention and it is powerful enough that participants can fail to be aware of random objects appearing on a screen or a gorilla walking through a basketball game [@simons1999gorillas; @mack2003inattentional]. Therefore if information is mapped to graphical elements that are so low on hierarchy they can be ignored, they might as well not be there at all. The people who don't plot uncertainty because they think its unimportant and the people who relegate uncertainty to the lowest ranking aesthetic on the list are both saying the same thing, "I think the uncertainty is unimportant". Including uncertainty is worth very little if no attention is left to see it. This does not mean we cannot put a *large* amount of information in a graphic. Glyph maps can be used to depict the multi-dimensional information in spatial-temporal data by mapping line plots to map locations, but we still need to decide what information is important. The map can either present trends in the global or local variance depending on whether or not the line plot is scaled globally or locally [@Wickham2012], however smaller details are almost impossible to convey. No matter how much information we try and put in a plot, there will always be only a handful of key takeaways.

>>>>

# Moved over notes
- @Blenkinsop2000 asked different questions for each uncertainty visualisation, indicating they had some awareness that the question has a significant impact on whether or not the visualisation should be used.
- @Gschwandtnei2016 When answering “What is the statistical probability (in %) that the interval has already ended at the marked point in time (red line)?” many answers gave the estimated probability to which the marked point in time falls into the interval. This is correct when asking for the probability to which the interval has already started at the marked position. However, in case of the probability to which the interval has already ended at the marked position, this answer is the inverted probability of what we were looking for. They believe this is their own mistake so they so they don’t present these results.

@Sanyal2009 users performed significantly well using Glyph-size when the task was to search for locations of least uncertainty. However, both Glyph-color and Surface-color performed better than Glyph- size when the task was to search for locations of high uncertainty. We did not expect to find a significant difference between the two search tasks since both the tasks were designed to find extremes in the data

@@

# Main section outtakes
These results provide some insights that applies to visualisation more broadly but nothing specific to uncertainty informaiton. It is clear that audiences primarily interact with visualisations through their "System 1" brain [@daniel2017thinking]. This aligns with other authors comments that visualisation is primarily about "gists" [Spiegelhalter2017]. In this sense, asking participants a question that requires them to shift to their "System 2" is entirely contrary to *why* we use visualisation. The exact reasons we describe visualisation to be so powerful, that they are able to communicate complicated information fast and efficiently *is a product of the system 1 brain* [@daniel2017thinking], however this means displaying information using visualisations have the same weakness' as system 1 thinking, and visualisation authors need to be aware of this. Asking participants to answer questions that require shifting to their system 2 brain is not in the spirit of visualisation, and does not reflect how visualisations will be used in practice, and will very likely result in incorrectly answered questions. 
There is a large number of examples where participants got "confused" but the problem is more likely that the question required participants to shift to their system 2 brain.

There is also the common view that uncertainty as a dimensionality problem. This perspective is not quite right and the literature seems to be about 50/50 on whether uncertainty is a psychological or dimensionality problem.  It is important to keep in mind that just because information is in a graphic, that does not mean it will be "seen". The phenomena of inattentional blindness shows that there is no perception without attention and it is powerful enough that participants can fail to be aware of random objects appearing on a screen or a gorilla walking through a basketball game [@simons1999gorillas; @mack2003inattentional]. Therefore if information is mapped to graphical elements that are so low on hierarchy they can be ignored, they might as well not be there at all. The people who don't plot uncertainty because they think its unimportant and the people who relegate uncertainty to the lowest ranking aesthetic on the list are both saying the same thing, "I think the uncertainty is unimportant". Including uncertainty is worth very little if no attention is left to see it. This does not mean we cannot put a *large* amount of information in a graphic. Glyph maps can be used to depict the multi-dimensional information in spatial-temporal data by mapping line plots to map locations, but we still need to decide what information is important. The map can either present trends in the global or local variance depending on whether or not the line plot is scaled globally or locally [@Wickham2012], however smaller details are almost impossible to convey. No matter how much information we try and put in a plot, there will always be only a handful of key takeaways.
